---
title: "Exploratory Data Analysis: Central Bank Independence and Inflation in Oil-Exporting Economies"
author: "Pavan Sundar - ECC3479 Research Project"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
    toc_depth: 3
    html_preview: false
---
```{r setup, include=FALSE}

# =========================
# PACKAGES
# =========================
if (!require("pacman")) install.packages("pacman")

pacman::p_load(
  tidyverse,
  modelsummary,
  scales,
  GGally,
  naniar,
  here
)

# =========================
# PROJECT ROOT
# =========================
project_root <- here::here()

# =========================
# OUTPUT DIRECTORY (FIGURES)
# =========================
out_dir <- file.path(project_root, "output", "eda_figures")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# =========================
# KNIT SETTINGS
# =========================
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  results = "asis",
  fig.width = 10,
  fig.height = 7,
  dev = "png",
  dpi = 300
)

# =========================
# DATA
# =========================
df <- read_csv(file.path(project_root, "data", "clean", "df_panel_final.csv"))

# =========================
# PLOT SAVE HELPER
# =========================
save_plot <- function(name, plot) {

  print(plot)

  ggplot2::ggsave(
    filename = file.path(out_dir, name),
    plot = plot,
    width = 10,
    height = 7,
    dpi = 300
  )
}

# =========================
# THEME
# =========================
theme_set(
  theme_classic(base_size = 13) +
    theme(
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.6),
      panel.grid = element_blank(),
      axis.line = element_blank(),
      axis.ticks = element_line(color = "black", linewidth = 0.4),
      axis.ticks.length = unit(2, "pt"),
      plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
      plot.subtitle = element_text(size = 11, colour = "grey30", hjust = 0.5),
      axis.title = element_text(size = 12),
      axis.text = element_text(size = 10),
      legend.position = "bottom",
      plot.margin = margin(12, 12, 12, 12)
    )
)
```

## 1. Data Structure and Cleaning Diagnostics

### 1.1 Data Structure
```{r data-structure}

glimpse(df)
```

Interpretation Notes:

Unit of observation is country-year panel data
Check whether panel is balanced or unbalanced
Identify obvious missing data or structural irregularities

### 1.2 Panel Structure

```{r, panel-structure}
cat("#### Table 1: Panel Dimensions\n\n")

datasummary_df(
  df %>% summarise(
    Observations = n(),
    Countries    = n_distinct(iso_code),
    Years        = n_distinct(year),
    Start_year   = min(year),
    End_year     = max(year)
  ))
```

Interpretation Notes:

A


```{r, coverage-years-per-country}
cat("#### Table 2: Coverage of years per country\n\n")

datasummary_df(
  df %>%
    distinct(iso_code, year) %>%
    count(iso_code, name = "Years_observed") %>%
    summarise(
      Min     = min(Years_observed),
      P25     = quantile(Years_observed, 0.25),
      Median  = median(Years_observed),
      Mean    = mean(Years_observed),
      P75     = quantile(Years_observed, 0.75),
      Max     = max(Years_observed)
    )
)
```

I
### 1.3 Missingness Structure
```{r, missing-data-structure}

save_plot(
  "01_missing_data_structure.png",
  df %>%
    miss_var_summary() %>%
    filter(pct_miss > 0) %>%
    mutate(pct_missing = pct_miss) %>%
    ggplot(aes(x = reorder(variable, pct_missing), y = pct_missing)) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Figure 1: Missing Data by Variable (%)",
      x = "",
      y = "Percent Missing"
    )
)
```

Interpretation Notes:

Are missing values random or systematic?
Which variables are most affected?
Implications for regression bias or sample selection

```{r, missingness-by-country}
cat("#### Table 3: Countries with highest missingness (key variables)\n\n")

datasummary_df(
  df %>%
    mutate(miss = rowSums(is.na(select(
      ., inflation_clean, cbi_extended, gdp_growth, m2_growth
    )))) %>%
    group_by(country_name) %>%
    summarise(
      Obs = n(),
      Avg_missing_vars = mean(miss),
      .groups = "drop"
    ) %>%
    arrange(desc(Avg_missing_vars)) %>%
    slice_head(n = 15),
)
```

Interpretation Notes:


## 2. Summary Statistics

### 2.1 Continuous Variables

```{r, summary-continuous}
cat("#### Table 4: Descriptive Statistics – Continuous Variables\n\n")

datasummary_skim(
  df %>% select(
    inflation_clean,
    inflation_rate,
    cbi_extended,
    unemployment_rate,
    trade_percent_gdp,
    gdp_growth,
    m2_growth,
    net_lending,
    oil_price_avg
  ),
  fmt = 2
)
```

Interpretation Notes:

Assess volatility in inflation and macro variables
Evaluate dispersion in CBI index
Identify potential outliers or crisis-driven skewness

### 2.2 Binary and Dummy Variables

```{r, summary-binary}

cat("#### Table 5: Exchange Rate Regime Composition\n\n")

datasummary_balance(
  ~ ex_regime_lbl,
  data = df %>%
    dplyr::mutate(
      ex_regime_lbl = factor(
        ex_regime,
        levels = c(0, 1),
        labels = c("Fixed", "Floating")
      )
    )
)

cat("#### Table 6: Net Oil Exporter Composition\n\n")

datasummary_balance(
  ~ is_net_exporter_lbl,
  data = df %>%
    dplyr::mutate(
      is_net_exporter_lbl = factor(
        is_net_exporter,
        levels = c(0, 1),
        labels = c("Net oil importer", "Net oil exporter")
      )
    )
)

cat("#### Table 7: Crisis Year Decomposition\n\n")

datasummary_balance(
  ~ crisis_year_lbl,
  data = df %>%
    dplyr::mutate(
      crisis_year_lbl = factor(
        crisis_year,
        levels = c(0, 1),
        labels = c("No crisis", "Crisis year")
      )
    )
)
```

Interpretation Notes:

What share of countries are floating vs pegged?
How frequent are crisis periods?
Is sample composition balanced or skewed?

## 3. Distributional Analysis

### 3.1 Inflation Distribution

```{r inflation-distribution}
cat("#### Table 8: Inflation Winsorisation Diagnostics\n\n")

datasummary_df(
  df %>% summarise(
    Min_raw      = min(inflation_rate, na.rm = TRUE),
    Max_raw      = max(inflation_rate, na.rm = TRUE),
    Min_clean    = min(inflation_clean, na.rm = TRUE),
    Max_clean    = max(inflation_clean, na.rm = TRUE),
    Pct_changed  = 100 * mean(inflation_clean != inflation_rate, na.rm = TRUE)
  ),
  fmt = 2
)

save_plot(
  "02_inflation_distribution.png",
  ggplot(df, aes(x = inflation_clean)) +
    geom_density(fill = "#2C3E50", alpha = 0.6) +
    labs(
      title = "Figure 2: Inflation Distribution (Winsorised)",
      x = "Inflation (%)",
      y = "Density"
    )
)
```

Interpretation Notes:

Skewness indicates presence of extreme inflation episodes
Justifies winsorisation strategy
Check for fat tails / crisis clustering

### 3.2 CBI Distribution

```{r, cbi-distribution}
save_plot(
  "03_cbi_distribution.png",
  ggplot(df, aes(x = cbi_extended)) +
    geom_histogram(bins = 30, fill = "#27AE60", color = "white") +
    labs(
      title = "Figure 3: Distribution of Central Bank Independence",
      x = "CBI Index",
      y = "Frequency"
    )
)
```

Interpretation Notes:

Is institutional reform gradual or stepwise?
Are countries clustered at low/high independence?


## 4. Macroeconomic Validity and Identification Structure

### 4.1 CBI vs Inflation (Main Identification Plot)

```{r, cbi-inflation-relationship}
save_plot(
  "04_cbi_inflation_relationship.png",
  ggplot(df, aes(x = cbi_extended, y = inflation_clean)) +
    geom_point(alpha = 0.15, color = "#2C3E50") +
    geom_smooth(method = "lm", color = "#E74C3C", linewidth = 1) +
    geom_smooth(method = "loess", color = "#3498DB", linetype = "dashed") +
    labs(
      title = "Figure 4: Central Bank Independence and Inflation",
      subtitle = "Linear vs Non-linear Relationship",
      x = "CBI Index",
      y = "Inflation (%)"
    )
)
```

Interpretation Notes:

Does higher CBI reduce inflation?
Linear vs non-linear structure
Evidence for credibility theory

### 4.2 Crisis Effects on Inflation

```{r, crisis-inflation}
save_plot(
  "05_crisis_inflation.png",
  df %>%
  mutate(crisis_year = ifelse(crisis_year == 1, "Crisis", "No Crisis")) %>%
  ggplot(aes(x = crisis_year, y = inflation_clean)) +
  geom_boxplot(fill = "#3498DB", alpha = 0.6) +
  labs(
    title = "Figure 5: Inflation: Crisis vs Non-Crisis Periods",
    x = "",
    y = "Inflation (%)"
  )
)
```

Interpretation Notes:

Crisis increases mean inflation?
Higher variance during crises?

### 4.3 Oil Shock Heterogeneity

```{r, oil-shock-heterogeneity}
save_plot(
  "06_oil_shock_heterogeneity.png",
  ggplot(
    df,
    aes(
      x = oil_price_avg,
      y = inflation_clean,
      color = as.factor(is_net_exporter)
    )
  ) +
    geom_smooth(method = "loess") +
    scale_color_discrete(
      labels = c(
        "0" = "Net Oil Importer",
        "1" = "Net Oil Exporter"
      )
    ) +
    labs(
      title = "Figure 6: Inflation Response to Oil Price Shocks",
      color = "Oil Trade Status",
      x = "Annual Brent Crude Price (US$/Barrel)",
      y = "Annual Inflation (%)"
    )
)
```

Interpretation Notes:

Do exporters hedge oil shocks better?
Asymmetric pass-through effects?

## 5. Institutional Heterogeneity and Simpson's Paradox

### 5.1 Exchange Rate Regime Interaction

```{r, ex-regime-interaction}
save_plot(
  "07_ex_regime_interaction.png",
  df %>%
    dplyr::filter(!is.na(ex_regime)) %>%   # 2️⃣ remove NA regime observations
    ggplot(aes(
      x = cbi_extended,
      y = inflation_clean
    )) +
      geom_point(alpha = 0.1) +
      geom_smooth(method = "lm", color = "#E74C3C") +
      facet_wrap(
        ~ ex_regime,
        labeller = as_labeller(c(
          `0` = "Fixed",
          `1` = "Floating"
        ))
      ) +
      labs(
        title = "Figure 7: CBI Effect by Exchange Rate Regime",
        x = "CBI Index",
        y = "Annual Inflation % change (Winsorised)"
      )
)
```

Interpretation Notes:

Does regime alter CBI effectiveness?
Evidence of Simpsons paradox?

### 5.2 Inflation Persistence

```{r, inflation-persistence}

cat("#### Table 9: Inflation Persistence Diagnostics\n\n")

datasummary_df(
  df %>% summarise(
    Correlation = cor(lag_inflation, inflation_clean, use = "complete.obs"),
    N = sum(complete.cases(lag_inflation, inflation_clean))
  ),
  fmt = 3,
)

save_plot(
  "08_inflation_persistence.png",
  df %>%
    mutate(
      lag_inflation_clean = pmin(
        pmax(lag_inflation, -5),
        100
      )
    ) %>%
    ggplot(
      aes(
        x = lag_inflation_clean,
        y = inflation_clean
      )
    ) +
      geom_point(alpha = 0.15) +
      geom_smooth(method = "lm", color = "red") +
      geom_abline(
        slope = 1,
        intercept = 0,
        linetype = "dashed"
      ) +
      labs(
        title = "Figure 8: Inflation Persistence (Winsorised)",
        x = "Lagged Inflation (% change, winsorised)",
        y = "Annual Inflation % change (winsorised)"
      )
)

cat("#### Table 10: Inflation persistence by Crisis Status\n\n")

datasummary_df(
  df %>%
    filter(complete.cases(lag_inflation, inflation_clean)) %>%
    group_by(crisis_year) %>%
    summarise(
      Correlation = cor(lag_inflation, inflation_clean),
      N = n(),
      .groups = "drop"
    ),
  fmt = 3
)
```
Interpretation Notes:

Strong persistence suggests dynamic panel model
Justifies lagged dependent variable


## 5.3. Correlation Structures

```{r, correlation-structure}

cat("#### Table 11: Correlation matrix — Key Variables\n\n")

datasummary_correlation(
  df %>% select(
    inflation_clean,
    cbi_extended,
    unemployment_rate,
    trade_percent_gdp,
    gdp_growth,
    m2_growth,
    oil_price_avg
  ),
  fmt = 2
)

save_plot(
  "09_correlation_structure.png",
  df %>%
    select(
      cbi_extended,
      inflation_clean,
      unemployment_rate,
      trade_percent_gdp,
      gdp_growth,
      m2_growth,
      oil_price_avg
    ) %>%
    ggcorr(
      label = TRUE,
      label_round = 2,
      layout.exp = 2        # ✅ KEY FIX: expands spacing
    ) +
    labs(
      title = "Figure 9: Correlation Structure of Key Macroeconomic Variables"
    ) +
    theme(
      plot.margin = margin(12, 12, 12, 24)  # small extra breathing room
    )
)
```

Interpretation Notes:

Multicollinearity risk?
Structural relationships between macro variables?

## 6. Key Conclusions and Summary

```{r key-findings}
cat("#### Table 11: Sample size under Common Restrictions\n\n")

datasummary_df(
  tibble::tibble(
    Sample = c(
      "Full Dataset",
      "Non-missing Inflation",
      "Non-missing Inflation & CBI",
      "Baseline Regression Sample"
    ),
    Observations = c(
      nrow(df),
      sum(!is.na(df$inflation_clean)),
      sum(!is.na(df$inflation_clean) & !is.na(df$cbi_extended)),
      sum(complete.cases(
        df$inflation_clean,
        df$cbi_extended,
        df$gdp_growth,
        df$m2_growth
      ))
    )
  )
)
```
Key Findings (TO BE WRITTEN IN PROSE):

Inflation exhibits [skewed / stable / volatile] distribution patterns
CBI is [clustered / uniform], suggesting institutional rigidity or reform waves
Preliminary evidence suggests a [negative / weak / nonlinear] relationship between CBI and inflation
Crisis periods significantly increase inflation volatility
Oil price shocks exhibit heterogeneous effects across exporter status
Inflation is strongly persistent, justifying dynamic modelling approaches
END OF EDA